%!TEX root = ../nwoods_thesis.tex

\section{Acknowledgements}

They say it takes a village to raise a child, and a grad student is just a toddler whose grape juice has fermented, so I have many people to thank.
I am indebted to too many advisors, colleagues, friends, and family members to name them all here, so I will highlight just a few for particular appreciation; the many omissions are solely for the purpose of keeping this document under 200 pages.

I must first thank the members of the Wisconsin CMS group, who have, understandably, played a central role in my life over the past five and a half years.
The professors, Wesley Smith, Sridhara Dasu, and Matt Herndon, are great physicists and master tacticians, and their clear-eyed advice has been extremely valuable throughout my time here.
Sascha Savin has been a great guide in the day-to-day business of being a physicist, and he has taught me an enormous amount about physics, data analysis, work habits, intracollaboration politics, and just about everything else.
The postdocs, Evan Friis, Maria Cepeda, Senka Duric, Isobel Ojalvo, Bhawna Gomber, and C{\'e}cile Caillol, are uniformly brilliant and endlessly helpful.
Everything would grind to a halt and nothing would ever get done if not for Dan Bradley, Ajit Mohapatra, Carl Vuosalo, and the rest of the computing team, to whom I should also apologize for all the times I completely broke everything.
Pam Klabbers and Ale{\v{s}} Svetek are founts of knowledge about the CMS trigger, and are fully responsible for any competence I appeared to display while on shift.
I would never have accomplished anything without the other UW CMS grad students.
Tom Perry, Aaron Levine, Laura Dodd, Devin Taylor, Tyler Ruggles, Nick Smith, Kenneth Long, James Buchanan, and Usama Hussain, and others with whom I overlapped more briefly, have been great friends and comrades-in-arms.
The intelligence, collective knowledge, and snark on continuous display in the group chat and in our office chats were invaluable to both my work and my mental health.

Ivan Furic and Darin Acosta helped me get my foot in the door during my REU at the University of Florida and taught me a lot when I had essentially no knowledge base to work from.

The friends I made in Madison and Geneva are far too numerous to mention individually, but they are wonderful people and I'm lucky to have them in my life.
A special shoutout goes to my flatmates, Laser Kaplan, Svet Hristozkov, Heather Russell, and Tessa Carver, for not just putting up with my idiosyncrasies but actively making home a place I wanted to go.

I never would have gone anywhere in life without my smart, funny, supportive family: my parents Anne and David, my siblings Julia and Peter, a small army of extended family, and a clowder of cats led by The Boss, Gadget.
I could not ask for a stronger foundation on which to build my future.
