%!TEX root = ../nwoods_thesis.tex

\chapter{Results}

A number of measurements and analyses fall under the umbrella of four-lepton physics, and results of the work presented in this thesis were presented in several journal articles and Physics Analysis Summary documents made public by the CMS collaboration.
The first CMS measurement of the {\ZZ} inclusive cross section at {13\TeV} used roughly half the 2015 dataset ({1.34\fbinv}) and was presented in Ref.~\cite{CMS:2015fnj}.
That analysis was expanded to use the whole {2.6\fbinv} collected in 2015, and to include the {\Zfourl} branching fraction measurement, as we reported in Ref.~\cite{Khachatryan:2016txa}.
With the full 2016 dataset, the {\ZZ} cross section and {\Zfourl} branching ratio were measured again to greater precision in Ref.~\cite{CMS:2017ruh}, which also included differential cross section measurements and aTGC limits.
An upcoming paper including these measurements will also include a combination of the 2015 and 2016 inclusive cross section measurements.
Differential cross sections with respect to jet-related observables, and searches for EWK {\ZZ} production and aTGCs, were reported in Ref.~\cite{CMS:2017dyw}, which will be split into two papers (one for the cross section measurements, one for the searches) for final journal publication.
The Higgs boson was studied in the four-lepton final state in Refs.~\cite{CMS:2016rqf,CMS:2016ilx,Sirunyan:2017exp}.
In the following, results for each topic are only shown for 2016 data, which are taken to supersede 2015 data.



\section{Four-Lepton Yield and Distributions}


\subsection{Full Spectrum}

The full four-lepton invariant mass spectrum is shown in Fig.~X.
The single-{\PZ} resonance can be seen below {100\GeV}, the Higgs resonance is visible---though unresolved with this binning---in the {\Zgs} region below $2m_\PZ$, where the doubly resonant {\ZZ} continuum begins.
Filled histograms representing the SM prediction from the Monte Carlo simulation and the estimated reducible background are stacked, with the total expected yield uncertainty overlayed on the prediction as a grey band.
Data points are shown as black dots on top with vertical lines to show their statistical uncertainty.
The dilepton invariant mass spectrum is shown for both {\Zgs} candidates in Fig.~X and for the {\Zgs} candidate closest to the nominal {\PZ} boson mass ($\PZ_1$) in Fig.~X.
Figure~X shows $m_{\PZ_2}$ plotted against $m_{\PZ_1}$ for data events representative of all four-lepton production.
Clusters of events with zero ({\Zfourl} and nonresonant $\Pa^\ast \Pa^\ast$ production), one ($\PH \to \PZ\PZ^\ast$ and nonresonant $\PZ\Pa^\ast$ production), and two (nonresonant {\ZZ} production) on-shell {\PZ} bosons can be clearly seen.


\subsection[\texorpdfstring{$\mathrm{Z} \to 4\ell$}{Z to 4l} Resonance]{$\mathbf{Z} \to \mathbf{4\ell}$ Resonance}

Expected and observed yields for events satisfying the {\Zfourl} selection criteria ($80 < m_{4\ell} < 100\GeV$) are shown in Table~X.
The invariant mass distribution of these events is shown in Fig.~X.
Figure~X shows $m_{\PZ_2}$ plotted against $m_{\PZ_1}$ for all data events consistent with {\Zfourl} production.
Predictions from Monte Carlo samples generally agree well with the data.


\subsection{Higgs Resonance}

Figure~X shows the four-lepton invariant mass around the Higgs resonance, which can be clearly seen above the SM continuum background, for events passing the Higgs selection ($m_{\PZ_2} > 12\GeV$, $\text{SIP}_\text{3D} < 4$ for all leptons).
Table~X shows the observed and expected yields in the mass range $118 < m_{4\ell} < 130\GeV$.
Here, SM continuum production---considered signal in all other parts of this analysis---is considered background.
Figures~X-Y show the $\PZ_1$ mass, the $\PZ_2$ mass, and the scatter plot of $m_{\PZ_2}$ against $m_{\PZ_1}$, in the same four-lepton mass region around the Higgs resonance.
Agreement between predictions and data is again good.


\subsection{ZZ Production}

Expected and observed yields for on-shell {\ZZ} events are shown in Table~X.
The corresponding four-lepton and {\PZ} boson candidate invariant masses are shown in Figs.~X and Y, respectively.
Figure~X shows the distribution of the number of jets in these events.
Figures~X and Y show the $m_{\Pj\Pj}$ and $\lvert \Delta \eta_{\Pj\Pj} \rvert$ distributions for tagging jet pairs in the dijet selection.
Agreement is, again, good.



\section{ZZ Fiducial and Total Cross Section}




\subsection[\texorpdfstring{$\mathrm{Z} \to 4\ell$}{Z to 4l} Branching  Fraction]{$\mathbf{Z} \to \mathbf{4\ell}$ Branching  Fraction}
A nice little measurement



\section{Differential Cross Sections}
With awesome plots



\section{Vector Boson Scattering}
Electroweak signal's electrostrength



\section{Anomalous Coupling Limits}
Winning at not finding things
